A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-01-29 analysis/paper/paper.Rmd
included a couple of references

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016; see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @lambeck1998; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@glorstad2020; @hafsten1957; @hafsten1983; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region. However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic. [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. The empirical expectation for this development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.

@@

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

To get at the relationship between sites and the contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located a significant distance from a common isobase perpendicular to this gradient. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase, to name but a few [e.g. @romundset2011; @romundset2019]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications (see above).

@@

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. Following from the expected deviance from the settlement patterns that characterise forager sites, these agricultural phases are thus treated seperately below. Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region. Due to the uncertain nature of this assertion these are omitted here.

@@

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site ranged from taking a few hours to a couple of days, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected. To what degree these do in fact represent contamination is discussed further below. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021]. With the transition from a predominantly hunter-gatherer lifeway to early agro-pastoralism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.

>>>>

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @lambeck1998; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@glorstad2020; @hafsten1957; @hafsten1983; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region. However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic. [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. The empirical expectation for the above outlined development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.

@@

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

To get at the relationship between sites and the contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b, @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located a significant distance from a common isobase perpendicular to this gradient. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. Following from the expected deviance from the settlement patterns that characterise forager sites, these agricultural phases are thus treated seperately below. Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region. Due to the uncertain and somewhat speculative nature of thi suggestion, these are omitted here.

@@

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site took up to a few hours, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021]. With the transition from a predominantly hunter-gatherer lifeway to early agro-pastoralism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.
